---
output: pdf_document 
---


```{r, echo=FALSE, eval=TRUE, message=FALSE}
  if (exists("lsg") == FALSE){ #Nur falls von RStudio aufgerufen
    lsg <- TRUE   #Wenn man die lsg ausgeben will dann lsg <- TRUE sonst 
  }
  if (exists("baseDir") == FALSE){ #Nur falls von RStudio aufgerufen
    baseDir = getwd()
  }
  ig <- function(name, scale = 1) {
     return (paste0("\\includegraphics[ scale =", scale,"]{", baseDir,"/", name, "}"))
  }
```


## Stan Sampling of the water on earth example
This is simular to the grid approximation, but know we do the sampling with Stan, otherwise the situation is the same.
We assume that the globe has been thrown $N = 9$ times and landed $W = 6$ times on water. 

a) Use a prior that reflect the fact that you have know for sure that more then 40 percent of the earth is covered by water but otherwise, you have no clue. Plot the posterior for all values of $p$. 

```{r, echo=lsg, eval=FALSE, code=readLines("~/Documents/GitHub/da/lab/water_world/world.stan"), collapse=TRUE}
#Stan Code
```

```{r, echo=lsg, eval=lsg, fig.height=4, warning=FALSE, error=FALSE, results='hide'}
library(cmdstanr)
# See https://github.com/oduerr/da/blob/master/lab/water_world/world.stan 
model = cmdstan_model('~/Documents/GitHub/da/lab/water_world/world.stan')
W = 6
N = 9
samples_cmd = model$sample(data = list(N=N, W=W), show_messages = FALSE)
samples_cmd = samples_cmd$draws('p', format = 'data.frame') #Extract parameter 'p'

#Comparison with the grid approximation
hist(samples_cmd$p, freq = FALSE, xlim=c(0,1))
lines(density(samples_cmd$p), col='green', lw=5)
```



b) What is the probability that there is more than 80 precent water on earth? Estimate again from the samples. Use samples from the posterior as in in c).

```{r, echo=lsg, eval=lsg, fig.height=4}
  mean(samples_cmd$p > 0.8)
```

