---
  output:
    github_document
---

# Excercise:

This notebook compares a non-hierarchical model and a hierarchical model to predict the scores of English premier league matches.

### Background

This notebook is a tutorial on Hierarchical modelling using Stan which is adopted from <https://github.com/MaggieLieu/STAN_tutorials/blob/master/Hierarchical/Hierarchical.Rmd>. The model is also described in the paper: <https://discovery.ucl.ac.uk/id/eprint/16040/1/16040.pdf>

Modifications include:

-   Removing of the pooled model

-   Removing the fixed prior for home advantage (which has a std of about 1e-4)

    -   However this seems to require that as in the paper we need to use a sum-to-zero constraint for the defense and offense. This makes sense intuitively, since the teams play against each other and therefore these capabilities are relative to each other. For more information on this see the paper above and the references therein.

```{r}
lsg = TRUE #Set to FALSE before submitting
require(rstan)
set.seed(1) #set seed 
```

### Reading the data

First we read in the data

```{r}
data = read.csv('https://raw.githubusercontent.com/MaggieLieu/STAN_tutorials/master/Hierarchical/premiereleague.csv',col.names = c('Home','score1', 'score2', 'Away'), stringsAsFactors = FALSE)
```

```{r}
head(data)
```

```{r}
ng = nrow(data)
cat('we have G =', ng, 'games \n')
```

```{r}
nt = length(unique(data$Home))
cat('We have T = ', nt, 'teams \n')
```

### Data Preparation

Now convert team names for each match into numbers (for the lookup)

```{r}
teams = unique(data$Home)
ht = unlist(sapply(1:ng, function(g) which(teams == data$Home[g])))
at = unlist(sapply(1:ng, function(g) which(teams == data$Away[g])))
```

```{r}
# we will save the last np games to predict
np=200
ngob = ng-np #number of games to fit
my_data = list(
  nt = nt, 
  ng = ngob,
  ht = ht[1:ngob], 
  at = at[1:ngob], 
  s1 = data$score1[1:ngob],
  s2 = data$score2[1:ngob],
  np = np,
  htnew = ht[(ngob+1):ng],
  atnew = at[(ngob+1):ng]
)
```

### **Traditional (non-hierachical) method**

We will assume that the goals scored come from a poisson distribution

-   s1 \~ Poisson(theta_g1) #game g score by home team

-   s2 \~ Poisson(theta_g2) #game g score by away team

Assuming a log-linear random effect model

-   log(theta_g1) = home + att_ht + def_at

-   log(theta_g2) = att_at + def_ht

Where home is a constant for the advantage for the team hosting the game att and def are the attack and defense abilities of the teams where the indices at,ht correspond to the t=1-20 teams.

```{r, results="hide"}
nhfit = stan(file = 'non_hier_model.stan', data = my_data)
```

#### Plot the predicted scores of the lastmatches

```{r, non_hier_pred_scores}
nhparams = extract(nhfit)
pred_scores = c(colMeans(nhparams$s1new),colMeans(nhparams$s2new))
true_scores = c(data$score1[(ngob+1):ng],data$score2[(ngob+1):ng] )
plot(true_scores, pred_scores, xlim=c(0,5), ylim=c(0,5), pch=20, ylab='predicted scores', xlab='true scores')
abline(a=0,  b=1, lty='dashed')

sqrt(mean((pred_scores - true_scores)^2))
cor(pred_scores, true_scores)
```

```{r, eval=lsg, show=lsg}
get_score = function(nhparams, data){
  s1_pred_score = round(colMeans(nhparams$s1new),0)
  s2_pred_score = round(colMeans(nhparams$s2new),0)
  points = 0
  for (i in 1:length(s1_pred_score)){
      pred_scorediff = mean(nhparams$s1new[,i] > nhparams$s2new[,i])
      if (s1_pred_score[i] == data$score1[ngob+i]
          && s2_pred_score[i] == data$score2[ngob+i]){
        points = points + 3
      } else {
        if (abs(pred_scorediff) < 0.5){ #Call it a draw
          if (data$score1[ngob+i] == data$score2[ngob+i]) {
            points = points + 1
          }
        } else {#No draw
           if(sign(pred_scorediff) == sign(data$score1[ngob+i] - data$score2[ngob+i])){
             points = points + 1
           }
        }
      } 
  }
  return (points)
}
get_score(nhparams, data)
```

#### Defense / Attack Non-Hierachical

We can also look at the attack/defense of the teams:

```{r attack_defense_non_hier}
attack = colMeans(nhparams$att)
defense = colMeans(nhparams$def)
plot(attack,defense,xlim=c(-0.4,1.1))
abline(h=0)
abline(v=0)
text(attack,defense, labels=teams, cex=0.7, pos=4)
```

Home Advantage

```{r home_adv_non_hier, show=lsg, eval=lsg}
home_adv = extract(nhfit)$home
hist(home_adv, 100, freq = FALSE, main=paste0("Home Adv, mean: ", round(mean(home_adv),2)))
lines(density(home_adv))
```

# Hierarchical model

In a hierarchical model, the parameters of interest, in our case the attack and defense ability are drawn from the population distribution.

-   att[t] \~ normal(mu_att, tau_att)

-   def[t] \~ normal(mu_def, tau_def)

Instead we define priors on the population, known as the hyperpriors.

```{r, results="hide"}
hfit = stan(file = 'hier_model.stan', data = my_data)
```

## Prediction of Hierarchical Model

```{r, hpreds, fig.height=5, fig.width=5}
hparams = extract(hfit)
pred_scores = c(colMeans(hparams$s1new),colMeans(hparams$s2new))
pred_errors = c(sapply(1:np, function(x) sd(hparams$s1new[,x])),sapply(1:np, function(x) sd(hparams$s1new[,x])))
true_scores = c(data$score1[(ngob+1):ng],data$score2[(ngob+1):ng] )
plot(true_scores, pred_scores, xlim=c(0,5), ylim=c(0,5), pch=20, ylab='predicted scores', xlab='true scores')
abline(a=0,  b=1, lty='dashed')
arrows(true_scores, pred_scores+pred_errors, true_scores, pred_scores-pred_errors, length = 0.05, angle = 90, code = 3, rgb(0,0,0,0.3))
sqrt(mean((pred_scores - true_scores)^2))
cor(pred_scores, true_scores)
```

```{r, eval=lsg, show=lsg}
 get_score(nhparams, data)
```

```{r attack_defense_hier}
attack = colMeans(hparams$att)
attacksd = sapply(1:nt, function(x) sd(hparams$att[,x]))
defense = colMeans(hparams$def)
defensesd = sapply(1:nt, function(x) sd(hparams$def[,x]))

plot(attack,defense, xlim=c(-0.4,1), ylim=c(-0.45,0.3), pch=20)
arrows(attack-attacksd, defense, attack+attacksd, defense, code=3, angle = 90, length = 0.04, col=rgb(0,0,0,0.2))
arrows(attack, defense-defensesd, attack, defense+defensesd, code=3, angle = 90, length = 0.04,col=rgb(0,0,0,0.2))
#abline(h=0)
#abline(v=0)
text(attack,defense, labels=teams, cex=0.7, adj=c(-0.05,-0.8) )
mean(attack)
mean(defense)
```

#### Home Advantage h-model

```{r home_adv_hier, show=lsg, eval=lsg}
home_adv = extract(hfit)$home
hist(home_adv, 100, freq = FALSE, main=paste0("Home Adv, mean: ", round(mean(home_adv),2)))
lines(density(home_adv))
```
