---
title: "Alk_in_Marathon_ETT"
author: "Tensorchiefs"
date: "22/2/22"
output:
  pdf_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
if (!exists("lsg")){
  lsg = TRUE
}
```

## Artifical Example ETT (Marathon and Drinks)
The aim of the exercise is to understand how to estimate the Effect of Treatment on the Treated. We idea is that you try to understand the code.

We have the following example of a causal model on the finishing time $Y$ in Marathon. This is strongly positively influenced if people go to a training or not $X=0,1$. Further, we assume that one gets slower the more drinks one has on average per day ($Z=0,1,..$). If on goes to the training or not depends also on $Z$ the number of drinks day. Altogether, we have the following SCM. 

### Definition of the SCM
```{r, eval=FALSE, show=TRUE}
    Z <- U$Z, 
    X <- ifelse(Z > 3, 0 , 1) * U$X, # x = f(z, Ux)
    Y <- -2*X + 0.1*Z*X + 0.5*Z^2 + U$Y # y = f(x,z, Uy)
```

Which produces the following graph:
```{r, eval=TRUE, show=FALSE, fig.height=3, fig.width=3}
  library(dagitty)
  d = dagitty('dag{
    X [pos="1,1"]
    Ux [pos="1,0.5"]
    Z [pos="2,0"]
    Uz [pos="2,-0.5"]
    Y [pos="3,1"]
    Uy [pos="3,0.5"]
    X <- Z -> Y 
    X -> Y
    Ux -> X
    Uz -> Z
    Uy -> Y
  }'
  )
  plot(d)
```


Starting from the SCM and observational data, we determine by simulation the factual $Y_{X=1}$ and counterfactual $Y_{X=0}$ for individuals who went the training $X=1$ and compute so the treatment effect on the treated:
$$
  \text{ETT} = E(Y_{X=1}-Y_{X=0}|X=1)
$$

### Simulation of the observed data

```{r, eval=TRUE, show=FALSE}
# Simulating the ETT of Eq. 4.18 see Pearl Primer 106
#set.seed(41)
n = 50000
###########

U = data.frame(
  Z = rbinom(n = n, size=10, p=0.3), #Number of Drinks  
  X = sample(c(1, 0), size = n, replace = TRUE, prob = c(0.9, 0.1)), 
  Y = rnorm(n,0,1)
)

#Taken from the SCM above
D = data.frame(
   Z <- U$Z, 
   X <- ifelse(Z > 3, 0 , 1) * U$X, # x = f(z, Ux)
   Y <- -2*X + 0.1*Z*X + 0.5*Z^2 + U$Y # y = f(x,z, Uy)
)
names(D) = names(U)
```

### Step 1 (Abduction) 
Use the evidence (here $X=1$) to determine the corresponding U. More precisely, we just store the U's and the data from the examples which have $X=1$. 
```{r}
#We condition on X==1
U_X1 = U[D$X == 1,]
D_X1 = D[D$X == 1,]

dim(D_X1)
```
### Step 2 (Action)
Adapt the SCM to the intervention. 

```{r}
get_adapted_data = function(U, D, X_inter){
  ret = data.frame(
    #Data Generating Process in the adapted model with intervened X
    Z <- U$Z, 
    X <- X_inter,
    #X <- ifelse(Z > 3, 0 , 1) * U$X,
    #Y <- -2*X + 0.5*Z + U$Y
    Y <- -2*X + 0.1*Z*X + 0.5*Z^2 + U$Y
  )
  names(ret) = names(D)
  return(ret)
}
```

### Step 3 (Prediction)
Get (counter)factuals outcomes $Y_{X=1}$ and $Y_{X=0}$ under treatment $X=1$ and $X=0$
```{r}
Y_1 = get_adapted_data(U_X1, D_X1, X_inter=1)$Y
sum(Y_1 != D_X1$Y) #0 Consistency 
Y_0 = get_adapted_data(U_X1, D_X1, X_inter=0)$Y
```

### Estimation of the ETT
```{r}
(ETT = mean(Y_1 - Y_0))
sd(Y_1 - Y_0)
```

## Estimatation from Obersvational data only
With Z being the adjustment set satisfying the backdoor criterion, we can estimate ETT via:

$$
\text{ETT} = E(Y_{X=1}-Y_{X=0}|X=1) = E(Y | X=1) - \sum_{z'} E(Y | X=0, Z=z') \cdot P(Z=z'|X=1)
$$

Details of the calculation. 

1. We estimate the probability $P(Z=z'|X=1)$ by counting the number of individuals drinking at level $n_{z'|X=1}$ compared to $n_{x=1} = \sum_{z'} n_{z'|X=1}$. If we do not observe a level $h'$ the respective probability is estimated to 0. 

2. The quantity $E(Y | X=0, H=h') = \sum_{y'} y' \cdot P(y'|X=0, H=h')$ is estimated by the average effect of those being $X=0$ and in level $H=h'$. If there is no individual with $X=0, H=h'$ i.e. $P(y'|X=0, H=h')=0$ with the effect that $E(Y | X=0, H=h')=0$. 

```{r}
  D_X1 = D[D$X == 1,]
  dim(D_X1)
  D_X0 = D[D$X == 0,]
  dim(D_X0)
  
  #First Term
  (T1 = mean(D_X1$Y))
  
  #Second Term
  z_1 = unique(D_X0$Z) # Ocurring z-levels in (X=0, Z=z)
  z_2 = unique(D_X1$Z) # Ocurring z-levels in (X=1, Z=z)
  all_z = sort(union(z_1, z_2))
  E_X0 = rep(0, length(all_z))
  P_X1 = rep(0, length(all_z))
  for (i in 1:length(all_z)){
    z = all_z[i]
    if (z %in% z_1) E_X0[i] = mean(D_X0[D_X0$Z == z,]$Y) 
    if (z %in% z_2) P_X1[i] = sum(D_X1$Z == z) 
  }
  P_X1 = P_X1 / sum(P_X1)
  tibble::tibble(all_z, E_X0, P_X1)
  T1 - sum(E_X0*P_X1)
```








