---
output:
  pdf_document:
    highlight: pygments
    includes:
      in_header: header_lsg.tex
---

```{r, echo=FALSE, eval=TRUE, message=FALSE}
  if (exists("lsg") == FALSE){ #Nur falls von RStudio aufgerufen
    lsg <- TRUE   #Wenn man die lsg ausgeben will dann lsg <- TRUE sonst 
  }
  if (exists("baseDir") == FALSE){ #Nur falls von RStudio aufgerufen
    baseDir = getwd()
  }
  ig <- function(name, scale = 1) {
     return (paste0("\\includegraphics[ scale =", scale,"]{", baseDir,"/", name, "}"))
  }
```

## Robust Regression

In this exercise, we investigate the beneficial use of t-distribution for robust regression.

Use the following code to create some data, including two outlieres.

```{r, echo=TRUE, eval=TRUE}
  ######## Creating Data #####
s <- matrix(c(1, .6, 
              .6, 1), 
            nrow = 2, ncol = 2)
set.seed(3)
#Data w/o outliers
d = data.frame(MASS::mvrnorm(n = 100, mu = c(0,0), Sigma = s))
colnames(d) = c('y','x')
#sorting
idx = sort(d$x, index.return = TRUE)$ix
d = d[idx,1:2]
o <- d

dat_no_outlier = list(
  N = nrow(o),
  x = o$x,
  y = o$y
)
o[c(1:2), 1] <- c(5, 4.5)
dat = list(
  N = nrow(o),
  x = o$x,
  y = o$y
)
plot(y ~ x, dat, main='data with 2 outliers')
```

a) Fit a linear model to the data (with and w/o outlier). For the slope and intercept assume normal priors centred around zero with spread of 10. For sigma assume a half normal distribution with spread of 10. Compare the posteriors for the slope a.


```{r, echo=lsg, eval=lsg, warning=FALSE,  results="hide"}
  compiled_model <- cmdstan_model('/Users/oli/Documents/GitHub/da/lab/robust_regression/1d_lin.stan')
  fit_no <- compiled_model$sample(dat_no_outlier)
  fit_no
  fit <- compiled_model$sample(dat)
  fit
```

```{r, echo=lsg, eval=lsg, warning=FALSE,  results="hide"}
  compiled_model <- cmdstan_model('/Users/oli/Documents/GitHub/da/lab/robust_regression/1d_lin.stan')
  fit_no <- compiled_model$sample(dat_no_outlier)
  fit_no
  fit <- compiled_model$sample(dat)
  fit
```

```{r, echo=lsg, eval=lsg, warning=FALSE}
a = fit$draws('a', format = 'df')$a
b = fit$draws('b', format = 'df')$b
a_no = fit_no$draws('a', format = 'df')$a
b_no = fit_no$draws('b', format = 'df')$b
plot(density(a_no), col='green', xlim=c(-1,1))
lines(density(a), col='red')  
```

b) Plot the data and include $E(y|x) = \bar{a} \cdot x + \bar{b}$ , where $\bar{a}$ and $\bar{b}$ are the respective posterior means. 

```{r, echo=lsg, eval=lsg, warning=FALSE}
  plot(y ~ x, dat, main='Non-Robust regressions')  
  
  gauss_a_no = mean(a_no)
  gauss_b_no = mean(b_no)
  curve(gauss_a_no*x + gauss_b_no, add = TRUE, lty=2, lwd=2, col='green')
  
  gauss_a = mean(a)
  gauss_b = mean(b)
  curve(gauss_a*x + gauss_b, add = TRUE, lty=2, lwd=2, col='red')
```

c) Robust regression: Modify the stan code above to output a t-distribution instead of a Gaussian. See e.g. https://mc-stan.org/docs/2_18/functions-reference/student-t-distribution.html how to parametrize a t-distribution. Ensure that the parameter is $\nu \ge 1$. Plot $E(y|x) = \bar{a} \cdot x + \bar{b}$. Fit the robust to the data with the outliers and plot it.   


```{r, echo=lsg, eval=lsg, warning=FALSE,  results="hide"}
  robust_model <- cmdstan_model('/Users/oli/Documents/GitHub/da/lab/robust_regression/1d_lin_robust.stan')  
  fit_robust <- robust_model$sample(dat)
```


```{r, echo=lsg, eval=lsg, warning=FALSE,  results="hide"}
  plot(y ~ x, dat, main='Robust regressions')  
  
  gauss_a_no = mean(a_no)
  gauss_b_no = mean(b_no)
  curve(gauss_a_no*x + gauss_b_no, add = TRUE, lty=2, lwd=2, col='green')
  
  gauss_a = mean(a)
  gauss_b = mean(b)
  curve(gauss_a*x + gauss_b, add = TRUE, lty=2, lwd=2, col='red')
  
  a = fit_robust$draws('a', format = 'df')$a
  b = fit_robust$draws('b', format = 'df')$b
  robust_a = mean(a)
  robust_b = mean(b)
  curve(robust_a*x + robust_b, add = TRUE, lty=2, lwd=2, col='blue')
  
```

d) Preform an estimation to a leave on out cross validation (PSIS-LOO), using the loo function from the loo package.

```{r, echo=lsg, eval=lsg, warning=FALSE,  results="hide"}
  m <-    cmdstan_model('/Users/oli/Documents/GitHub/da/lab/robust_regression/1d_lin.stan')
  fit <- m$sample(dat)
  d = loo::loo(fit$draws("log_lik"))
  print("Data with outliers non-robust")
  print(d$estimates)

  fit <- m$sample(dat_no_outlier)
  d = loo::loo(fit$draws("log_lik"))
  print("Data w/o outliers non-robust")
  print(d$estimates)
  
   m <-    cmdstan_model('/Users/oli/Documents/GitHub/da/lab/robust_regression/1d_lin_robust.stan')
  fit <- m$sample(dat)
  d = loo::loo(fit$draws("log_lik"))
  print("Data with outliers robust")
  print(d$estimates)

  fit <- m$sample(dat_no_outlier)
  d = loo::loo(fit$draws("log_lik"))
  print("Data w/o outliers robust")
  print(d$estimates)
```


