---
output:
  pdf_document:
    highlight: pygments
    includes:
      in_header: header_lsg.tex
---

```{r, echo=FALSE, eval=TRUE, message=FALSE}
  if (exists("lsg") == FALSE){ #Nur falls von RStudio aufgerufen
    lsg <- TRUE   #Wenn man die lsg ausgeben will dann lsg <- TRUE sonst 
  }
  if (exists("baseDir") == FALSE){ #Nur falls von RStudio aufgerufen
    baseDir = getwd()
  }
  ig <- function(name, scale = 1) {
     return (paste0("\\includegraphics[ scale =", scale,"]{", baseDir,"/", name, "}"))
  }
```

## Linear Regression with Stan and Maximum Likelihood

In this excerise, we will use Stan to do a linear regression. The aim is to use Stan to describe the problem in the setting of the probabilistic programming language Stan. 

a) We simulate 50 data points $(x,y)$ under the assumptions from linear regression $y \sim Norm(\mu=2.0 \cdot x + 1.0, \sigma=1)$. Let $x$ be in the range from -2 to 2. Hint: use the function $rnorm$. NOte that the data is generated with a fixed x and a random y.

```{r, echo=lsg, eval=lsg}
  set.seed(1)
  N = 50
  x = seq(-2,2, length.out = N)
  y = rnorm(N, mean = 2 * x + 1, sd = 1)
```

b)  Fit the maximum likelihood solution with the R command `lm`. Plot the Data and $\mu_{ML} = E(y|x)$

```{r, echo=lsg, eval=lsg, warning=FALSE}
  maxl = lm(y ~ x)
  d = coef(maxl)
  a = d[2]
  b = d[1]
  curve(a*x+b, from = -5, to=5)
  points(x,y)
  a
  b
```

c) Formulate the problem in Stan. In case you are stuck, the stan-file can be found at <https://github.com/oduerr/da/blob/master/lab/lr_1/lr_maxlike.stan>. After defining the model, you can get the maximum likelihood solution by using the optimization algorithm from Stan via:

```{r, eval=lsg, echo=TRUE, warning=FALSE}
library(cmdstanr)
#Reading the model definition 
m_rcmdstan <- cmdstan_model('~/Documents/GitHub/da/lab/lr_1/lr_maxlike.stan')
data_list = list(N=N,x=x, y=y)
# Fitting the Maximum Likelihood Solution using Optimization
res = m_rcmdstan$optimize(data = data_list)
res$mle() #The maximum likelihood solution
```

You task is to create the Stan model `lr_maxlike.stan`.
```{r, echo=lsg, eval=FALSE, code=readLines('~/Documents/GitHub/da/lab/lr_1/lr_maxlike.stan'), collapse=TRUE}
# print the Stan model
m_rcmdstan
```
