---
output:
  pdf_document:
    highlight: pygments
    includes:
      in_header: header_lsg.tex
---

```{r, echo=FALSE, eval=TRUE, message=FALSE}
  if (exists("lsg") == FALSE){ #Nur falls von RStudio aufgerufen
    lsg <- TRUE   #Wenn man die lsg ausgeben will dann lsg <- TRUE sonst 
  }
  if (exists("baseDir") == FALSE){ #Nur falls von RStudio aufgerufen
    baseDir = getwd()
  }
  ig <- function(name, scale = 1) {
     return (paste0("\\includegraphics[ scale =", scale,"]{", baseDir,"/", name, "}"))
  }
```

## Regression of CO2 data

a) The following code will load the data set `co2` and plot it. The data set contains the monthly mean CO2 concentrations (in parts per million by volume) collected at the Mauna Loa Observatory in Hawaii, between 1959 and 1997. The data set contains two columns: the first column is the time in months (starting with 1 and ending with years*12), and the second column is the CO2 concentration.

```{r, echo=TRUE, eval=TRUE}
 # Loading the CO2 dataset
 data("co2", package = "datasets")
 co2_df <- data.frame(time = 1:length(co2), co2)
 head(co2_df)
 plot(co2, pch='.', main="CO2 concentration", xlab="Time (month)", ylab="CO2 concentration")
```

b) Use the following code for a linear regression. What is the average increase of CO2 per year?
```{r, echo=lsg, eval=lsg}
  maxl = lm(co2 ~ time, data = co2_df)
  d = coef(maxl)
  a = d[2]
  b = d[1]
  plot(co2_df, pch='.', main="CO2 concentration", xlab="Time (month number)", ylab="CO2 concentration")
  abline(maxl, col="red")
  maxl #Yearl 0.109 * 12
```

\lsg{"The average increase of CO2 per year is 0.109 * 12 = ppm."}


c) Formulate the problem in Stan. Load the model co2.stan from `` understand the code and run it. You will run into problems, provide good starting values for the optimization.

```{r, eval=lsg, echo=TRUE, warning=FALSE}
# Compile the model
mod <- cmdstan_model("~/Documents/GitHub/da/lab/pp/co2.stan")
# Prepare data for Stan
stan_data <- list(N = nrow(co2_df), time = co2_df$time, CO2 = co2_df$co2)
# Use the optimizing function
mle_results <- mod$optimize(data = stan_data)
par = as.data.frame(t(mle_results$mle()))
par
```

d) Use the following code to plot your prediction. 

```{r, echo=lsg, eval=lsg}
# Assuming mle_results contains the output from the Stan optimization
alpha_mle <- par$alpha
b_mle <-  par$b
a_mle <- par$a
omega_mle <- par$omega
phi_mle <- par$phi

# Generate predictions
predicted_co2 <- par$alpha + par$b * co2_df$time + par$a * sin( par$omega * co2_df$time + par$phi)

# Add predictions to the data frame
plot(co2_df$time, co2_df$co2, pch='.', main="CO2 concentration", xlab="Time (month)", ylab="CO2 concentration", type = 'l', col='blue')
lines(1:468,predicted_co2)
```


e) You will see that the fit is not perfect. Try to improve the fit by changing the starting values. 

```{r, eval=lsg, echo=TRUE, warning=FALSE}
initial_omega <- 2 * pi / 12  # Initial guess for the frequency of by 1 to see  
initial_a <- 1  # Small amplitude, adjust based on visual inspection if needed
initial_phi <- 0  # No phase shift initially
initial_sigma <- sd(resid(maxl))
initial_alpha = coef(maxl)[1]
initial_b = coef(maxl)[2]


init = list( a = initial_a, omega = initial_omega, phi = initial_phi, sigma = initial_sigma, alpha = initial_alpha, b = initial_b, c=initial_c)

mle_results <- mod$optimize(iter=1000, data = stan_data, init = list(init))

par = as.data.frame(t(mle_results$mle()))
par
# Generate predictions
predicted_co2 <- par$alpha + par$b * co2_df$time +  par$a * sin( par$omega * co2_df$time + par$phi)

# Add predictions to the data frame
plot(co2_df$time, co2_df$co2, pch='.', main="CO2 concentration", xlab="Time (month)", ylab="CO2 concentration", type = 'l', col='blue')
lines(1:468,predicted_co2)
```