---
output:
  pdf_document:
    highlight: pygments
    includes:
      in_header: header_lsg.tex
---

```{r, echo=FALSE, eval=TRUE, message=FALSE}
  if (exists("lsg") == FALSE){ # Only if called from RStudio
    lsg <- TRUE   # Set lsg <- TRUE to output solutions, otherwise FALSE
  }
  if (exists("baseDir") == FALSE){ # Only if called from RStudio
    baseDir = getwd()
  }
  ig <- function(name, scale = 1) {
     return (paste0("\\includegraphics[ scale =", scale,"]{", baseDir,"/", name, "}"))
  }
```

## Logistic Regression with Stan for Challenger Data

In this exercise, we will use the Challenger O-ring data to model the probability of O-ring failure as a function of temperature using logistic regression in Stan. Since this is a simple model,
we can interpret the coefficients to understand the relationship between temperature and the probability of O-ring failure.

a) Load the Challenger O-ring data. This dataset includes temperatures and O-ring failure indicators from the Space Shuttle launches:

```{r, echo=TRUE, eval=TRUE}
  url <- "https://raw.githubusercontent.com/oduerr/da/master/data/challenger.txt"
  challenger <- read.csv(url, header = TRUE)
```

b) Analyze the data by plotting the temperature against O-ring failure status:

```{r, echo=lsg, eval=lsg, warning=FALSE}
  #Add jitter to the data to avoid overplotting
  plot(jitter(challenger$Temp, factor = 1), challenger$Failure, xlab = "Temperature (F)", ylab = "O-ring Failure", pch = 19)
```

You can fit the logistic regression model using the R command `glm` with family binomial to find the maximum likelihood estimates as follows:

```{r, echo=TRUE, eval=TRUE, warning=FALSE}
  glm_fit = glm(Failure ~ Temp, family = binomial, data = challenger)
  coef(glm_fit)
  logLik(glm_fit)
```

c) Formulate the logistic regression problem in Stan. The Stan code should model the probability of failure based on temperature. Use Stan's optimization algorithm to find the maximum likelihood solution via

```{r, eval=FALSE, echo=TRUE, warning=FALSE}
library(cmdstanr)
# Reading the model definition
model_path <- 'lr_logistic.stan'  # Adjust the path as necessary
stan_model <- cmdstan_model(model_path)
data_list <- list(N = nrow(challenger), Temp = challenger$Temp, Failure = challenger$Failure)
# Fitting the Maximum Likelihood Solution using Optimization
result <- stan_model$optimize(data = data_list)
result$mle # The maximum likelihood solution
```

Your task is to create the Stan model `lr_logistic.stan`. 
Here are some tips to 

```{r, eval=FALSE, echo=TRUE}
data {
  int<lower=0> N;                    
  array[N] real Temp;                
  array[N] int<lower=0, upper=1> Failure;   // failure status at each observation
}

parameters {
  real alpha;                         // intercept
  real beta;                          // coefficient for temperature
}

model {
  // Logistic function for probability calculation
  for (i in 1:N) {
    real p = 1 / (1 + exp(-(alpha + TODO Temp[i])));  // Logistic function
    Failure[i] ~ bernoulli(p);                        // Bernoulli distribution with probability p
  }
}
```

```{r, eval=lsg, echo=lsg, warning=FALSE}
library(cmdstanr)
# Reading the model definition
model_path <- '~/Documents/GitHub/da/lab/pp/lr_logistic.stan'  # Adjust the path as necessary
stan_model <- cmdstan_model(model_path)
data_list <- list(N = nrow(challenger), Temp = challenger$Temp, Failure = challenger$Failure)
# Fitting the Maximum Likelihood Solution using Optimization
result <- stan_model$optimize(data = data_list)
result$mle() # The maximum likelihood solution
```

e) The coefficients from the logistic regression model give us insights into how temperature affects the probability of O-ring failure. The coefficient for temperature (beta) indicates how the log-odds of failure change with a one-degree increase in temperature. That is

$$
\frac{odd(T \rightarrow T+1)}{odd(T)} = exp(\beta)
$$


Calculate the odds ratio (OR) for the temperature coefficient to interpret the effect of temperature on the probability of O-ring failure. Interpret the result.

```{r, echo=lsg, eval=lsg, warning=FALSE}
  # Calculate the odds ratio for the temperature coefficient
  beta = result$mle()[2]
  odds_ratio <- exp(beta)
  odds_ratio 
  #For each degree increase in temperature, the odds of O-ring failure needs to be multiplied by the odds  ratio of 0.79. Thus the risk of O-ring failure decreases with by 21% for each degree increase in temperature. This is independent of the current temperature.
```

g) Finally, plot the logistic regression curve to visualize the relationship between temperature and the probability of O-ring failure:

```{r, echo=lsg, eval=lsg, warning=FALSE}
  # Define the logistic function
  logistic <- function(x) {
    return(1 / (1 + exp(-x)))
  }
  
  # Plot the logistic regression curve
  x <- seq(min(challenger$Temp)*0.5, max(challenger$Temp)*1.5, length.out = 100)
  y <- logistic(result$mle()[1] + result$mle()[2] * x)
  plot(challenger$Temp, challenger$Failure, xlab = "Temperature (F)", ylab = "Prob. of O-ring Failure", pch = 19)
  lines(x, y, col = "red")
```
