---
  output:
    github_document
  params:
    lsg: FALSE 
---

# Excercise:

This notebook compares a non-hierarchical model and a hierarchical model to predict the scores of English premier league matches.

## Your tasks

Step through the notebook and try to understand it. Try to answer the following questions:

- Why the $log$ in $log(theta_g2) = att_at + def_ht$?
- For 15 games and for 200 games in the training set. Answer the following questions:
  - What are the effective number of parameters in the hierarchical model and non-hierarchical model (use the output of the `loo` function)?
  - What is the average score (for definition see below) and  
  
- What is the home advantage in the non-hierarchical model and the hierarchical model?


### Background

This notebook is a tutorial on Hierarchical modelling using Stan which is adopted from <https://github.com/MaggieLieu/STAN_tutorials/blob/master/Hierarchical/Hierarchical.Rmd>. The model is also described in the paper: <https://discovery.ucl.ac.uk/id/eprint/16040/1/16040.pdf>

Modifications include:

-   Removing of the pooled model

-   Removing the fixed prior for home advantage (which has a std of about 1e-4)

-   Nevertheless, it appears necessary, as mentioned in the paper, to implement a sum-to-zero restriction for both the defense and offense. This stems from the fact that teams compete against one another, making these abilities dependent on each other. For further details, please refer to the aforementioned paper and its associated references.

```{r}
lsg = TRUE #Set to FALSE before submitting
#require(rstan)
#set.seed(1) #set seed 
library(cmdstanr)
options(mc.cores = parallel::detectCores())
```

### Reading the data

First we read in the data

```{r}
data = read.csv('https://raw.githubusercontent.com/MaggieLieu/STAN_tutorials/master/Hierarchical/premiereleague.csv',col.names = c('Home','score1', 'score2', 'Away'), stringsAsFactors = FALSE)
```

```{r}
head(data)
```

```{r}
ng = nrow(data)
cat('we have G =', ng, 'games \n')
```

```{r}
nt = length(unique(data$Home))
cat('We have T = ', nt, 'teams \n')
```

### Data Preparation

Now convert team names for each match into numbers (for the lookup)

```{r}
#Some R-Magic to convert the team names to numbers, no need to understand this
teams = unique(data$Home)
ht = unlist(sapply(1:ng, function(g) which(teams == data$Home[g])))
at = unlist(sapply(1:ng, function(g) which(teams == data$Away[g])))
```

```{r}
# we will save the last np games to predict
np=300
ngob = ng-np #ngames obsered ngob = number of games to fit
print(paste0("Using the first ", ngob, " games to fit the model and ", np, " games to predict"))
my_data = list(
  nt = nt, 
  ng = ngob,
  ht = ht[1:ngob], 
  at = at[1:ngob], 
  s1 = data$score1[1:ngob],
  s2 = data$score2[1:ngob],
  np = np,
  htnew = ht[(ngob+1):ng],
  atnew = at[(ngob+1):ng]
)
```

### Traditional (non-hierachical) method

We will assume that the goals scored come from a poisson distribution

-   s1 \~ Poisson(theta_g1) #game g score by home team

-   s2 \~ Poisson(theta_g2) #game g score by away team

Assuming a log-linear random effect model

-   log(theta_g1) = home + att_ht + def_at

-   log(theta_g2) = att_at + def_ht

Where home is a constant for the advantage for the team hosting the game att and def are the attack and defense abilities of the teams where the indices at,ht correspond to the t=1-20 teams.

```{r, results="hide",  message=FALSE, warning=FALSE,  results="hide"}
library(here)
library(cmdstanr)
nhmodel <- cmdstan_model(here('lab/football1/non_hier_model.stan'))
nhfit = nhmodel$sample(data = my_data)
d = loo::loo(nhfit$draws("log_lik"))
print(d)
```

#### Non-Hierarchical model predicted scores of the unseen

```{r non_hier_pred_scores}
library(tidyverse)
s1new = nhfit$draws(format = "df", 's1new') %>% select(-.chain, -.iteration, -.draw)
s2new = nhfit$draws(format = "df", 's2new') %>% select(-.chain, -.iteration, -.draw)
pred_scores = c(colMeans(s1new),colMeans(s2new))
true_scores = c(data$score1[(ngob+1):ng],data$score2[(ngob+1):ng] )
plot(true_scores, pred_scores, xlim=c(0,5), ylim=c(0,5), pch=20, ylab='predicted scores', xlab='true scores')
abline(a=0,  b=1, lty='dashed')

sqrt(mean((pred_scores - true_scores)^2))
cor(pred_scores, true_scores)
```
Below is a function to calculate the score of the model. Using the following rules:

1. Exact Score Prediction:
	3 Points are awarded if both the predicted home and away scores match the actual scores exactly.
1.	Draw Prediction:
	1 Point is awarded if the predicted result is a draw (both teams score the same number of goals) and the actual result is also a draw.
1.	Correct Outcome Prediction:
	1 Point is awarded if the predicted outcome (win/loss) matches the actual outcome, even if the exact scores do not match.


```{r, eval=lsg, show=lsg}
get_score = function(s1new, s2new, data){
  s1_pred_score = round(colMeans(s1new),0)
  s2_pred_score = round(colMeans(s2new),0)
  points = 0
  for (i in 1:length(s1_pred_score)){
      pred_scorediff = s1_pred_score - s2_pred_score
      if (s1_pred_score[i] == data$score1[ngob+i]
          && s2_pred_score[i] == data$score2[ngob+i]){
        points = points + 3
      } else {
        if (s1_pred_score[i] == s2_pred_score[i]){ #Call it a draw
          if (data$score1[ngob+i] == data$score2[ngob+i]) {
            points = points + 1
          }
        } else {#No draw
           if(sign(pred_scorediff[i]) == sign(data$score1[ngob+i] - data$score2[ngob+i])){
             points = points + 1
           }
        }
      } 
  }
  return (points)
}
get_score(s1new, s2new, data)
```

#### Defense / Attack Non-Hierachical

We can also look at the attack/defense of the teams:

```{r attack_defense_non_hier}
attack = nhfit$draws(format = "df", 'att') %>% 
  select(-.chain, -.iteration, -.draw) %>% colMeans()
 
defense = nhfit$draws(format = "df", 'def') %>% 
  select(-.chain, -.iteration, -.draw) %>% colMeans()
plot(attack,defense,xlim=c(-0.4,1.1))
abline(h=0)
abline(v=0)
text(attack,defense, labels=teams, cex=0.7, pos=4)
```

### Your Excercise / Determine the home advantage

Plot the posterior of the variable `home`

```{r home_adv_non_hier, show=lsg, eval=lsg, echo=lsg}
library(tidybayes)
#home_adv = rstan::extract(nhfit)$home
#hist(home_adv, 100, freq = FALSE, main=paste0("Home Adv, mean: ", round(mean(home_adv),2)))
#lines(density(home_adv))

tidy_draws(nhfit) %>% 
  select(home) %>% 
  ggplot(aes(x=home)) + geom_density() 
```

# Hierarchical model

In a hierarchical model, the parameters of interest, in our case the attack and defense ability are drawn from the population distribution.

-   att[t] \~ normal(mu_att, tau_att)

-   def[t] \~ normal(mu_def, tau_def)

Instead we define priors on the population, known as the hyperpriors.

```{r hmodel, results="hide", message=FALSE, warning=FALSE,  results="hide"}
hmodel <- cmdstan_model(here('lab/football1/hier_model.stan')) 
hfit = hmodel$sample(data = my_data)
```

## Checking the chain

```{r }
hfit
bayesplot::mcmc_trace(hfit$draws(c("mu_att", "lp__")))
```

```{r }
loo::loo(hfit$draws("log_lik"))
loo::loo(nhfit$draws("log_lik"))
```

## Prediction of Hierarchical Model

```{r hpreds, fig.height=5, fig.width=5}
s1new = hfit$draws(format = "df", 's1new') %>% select(-.chain, -.iteration, -.draw)
s2new = hfit$draws(format = "df", 's2new') %>% select(-.chain, -.iteration, -.draw)
pred_scores = c(colMeans(s1new),colMeans(s2new))
true_scores = c(data$score1[(ngob+1):ng],data$score2[(ngob+1):ng] )
plot(true_scores, pred_scores, xlim=c(0,5), ylim=c(0,5), pch=20, ylab='predicted scores', xlab='true scores')
abline(a=0,  b=1, lty='dashed')

sqrt(mean((pred_scores - true_scores)^2))
cor(pred_scores, true_scores)
```

```{r, eval=lsg, show=lsg}
 get_score(s1new, s2new, data)
```

### Mean attack and dense abilities

```{r attack_defense_hier}
attack = hfit$draws(format = "df", 'att') %>% 
  select(-.chain, -.iteration, -.draw) %>% colMeans()
 
defense = hfit$draws(format = "df", 'def') %>% 
  select(-.chain, -.iteration, -.draw) %>% colMeans()
plot(attack,defense,xlim=c(-0.4,1.1))
abline(h=0)
abline(v=0)
text(attack,defense, labels=teams, cex=0.7, pos=4)
```

#### Home Advantage h-model you job!

```{r home_adv_hier, show=lsg, eval=lsg, echo=lsg}
tidy_draws(hfit) %>% 
  select(home) %>% 
  ggplot(aes(x=home)) + geom_density() 
```
